---
title: "tsdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tsdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(comp.stat.final.assignment)
```

## Introduction

The `comp.stat.final.assignment` package defines an S3 class named `tsdata` for representing univariate time series data. Each `tsdata` object stores a numeric vector of values and a corresponding vector of `Date` objects. This package provides methods for printing, summarizing, and plotting these objects.

## Creating a `tsdata` Object

You can create a `tsdata` object by using the `tsdata()` function, which takes two arguments: a numeric vector `data` and a `Date` vector `dates` of the same length.

```{r}
# Simulate 100 daily observations
set.seed(123)
data <- rnorm(100, mean = 10, sd = 2)
dates <- seq.Date(from = as.Date("2025-01-01"), by = "day", length.out = 100)

# Create a tsdata object
x <- tsdata(data = data, dates = dates)
```

## Printing the `tsdata` Object

The `print()` method gives a short summary of the object, including the number of observations and the time span:

```{r}
print(x)
```

## Summarizing the `tsdata` Object

The `summary()` method returns useful statistics such as the number of observations, mean, median, standard deviation, minimum, maximum, and the start/end dates:

```{r}
summary(x)
```

## Plotting the `tsdata` Object

The `plot()` method creates a time series plot using `ggplot2`, which makes it easy to visualize the time series.

```{r}
plot(x)
```
